---
title: "Formative assessment"
author: "Sriram Bapatla"
format: html
editor: visual
---

## Playing around

First, let's play around with the distribution, as I'm pretty unfamiliar with it overall. Let's start with the simpler case of $\xi=0$, meaning a pdf of $$f(x)= 1/\sigma\exp(-\frac{x-u}{\sigma}) $$ Let's view a couple distributions across a range of parameters for $u$ and $\sigma$.

```{r}
u = 2
x<-seq(u,5,0.1)
sigma = .2

plot(x,(1/sigma)*exp(-(x-u)/sigma),pch = 16)

```

This is just the exponential function when $\xi=0$. Now, let's see what it looks like when we add in the scaling factor.

```{r}

u = 0
x<-seq(u,5,0.1)
sigma = 1
xi1 = 4
xitests<- c(-3,-2,-1,-0.5,-0.2)

plot(x, (1/sigma)*(1 + xi*(x-u)/sigma)**(-1/xi-1),pch = 16)
for (i in 1:5) {
  points(x, (1/sigma)*(1 + xitests[i]*(x-u)/sigma)**(-1/xitests[i]-1),pch = 16)
  print(i)
}

```

Okay so now let's go to the problem

```{r}
u = 2
x<-seq(0.001,1,.01)
sigma = .2

##for xi = 0
cdf<- 1-exp(-1*(x-u)/sigma)
quantile_test<- u-sigma*log((1-x),exp(1))

plot(x,quantile_test,pch = 16)

```
Cool now the general case 

``` {r}

u = 2
x<-seq(0.001,1,.01)
sigma = .2
xi = 1
y<- (sigma/xi)*((1-y)**(-1*xi) - 1) + u

dom<-seq(u,5,0.1)
cdf<- 1- (1+xi*(dom-u)/sigma)**(-1/xi)

plot(x,y,pch = 16)
plot(dom,cdf,pch = 16)



```



